---
title: "Case Study 1: Intersectional Bias in Recidivism Prediction"
output: 
  github_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, cache = FALSE,
                      fig.height = 4)
```

This guide book completely reproduces Case Study 1 in the paper *seeBias: A
Comprehensive Tool for Assessing and Visualizing Fairness in AI Prediction
Models*, which evaluates bias across race and sex when predicting two-year
recidivism using the COMPAS data.

## Build prediction model

The COMPAS data is sourced from the [*fairmodels*](https://fairmodels.drwhy.ai/)
R package and also provided in the *seeBias* package.

```{r}
library(seeBias)
library(dplyr) # to manipulate data
library(tableone) # to describe final dataset
library(knitr) # to format tables
library(ggplot2) # to edit figures
library(ggpubr) # to arrange figures
data("compas")
knitr::kable(head(compas))
```

### Prepare data

Use subset of COMPAS data for White and Black individuals. Exclude race and 
sex from the prediction model.

```{r}
dat <- compas %>% filter(Ethnicity %in% c("Caucasian", "African_American"))
dat$Ethnicity <- ifelse(dat$Ethnicity == "African_American", "Black", "White")
```

The COMPAS dataset includes `r nrow(compas)` individuals. We focused on a subset
of `r nrow(dat)` White and Black individuals and examined bias across race
(`Ethnicity`) and sex (`Sex`).

```{r}
CreateTableOne(vars = names(dat), strata = "Two_yr_Recidivism", 
               data = dat, addOverall = TRUE) %>% 
  kableone(caption = "Supplementary Table S1. Characteristics of individuals analyzed in the prediction of two-year recidivism.")
```

### Build logistic regression model

In this example, all subjects in the final dataset are used for model training
and evaluation.

```{r,message=FALSE}
m <- dat %>% select(-Ethnicity, -Sex) %>% 
  glm(Two_yr_Recidivism ~ ., data = ., family = "binomial")
kable(
  cbind(exp(coef(m)), exp(confint(m))), digits = 2, 
  col.names = c("OR", "95% CI (lower bound)", "95% CI (upper bound)"), 
  caption = "Odds ratio (OR) and 95% confidence interval (CI) from the logistic regression model."
)
```

## Fairness evaluation

Construct the seeBias object, using race and gender as sensitive variables.

```{r}
# Extracted predicted risk and observations from test data.
# If not specified, the best threshold in ROC analysis is used.
x <- dat %>% select(Ethnicity, Sex) %>% evaluate_prediction_prob(
  y_pred = predict(m, newdata = dat, type = "response"), 
  y_obs = dat$Two_yr_Recidivism, y_pos = "1",
  sens_var = ., sens_var_ref = c("White", "Male")
)
```

### Compute fairness metrics

Table 1 in the manuscript:

```{r results='asis'}
summary(x)
```

### Visualise fairness metrics

By default, only the figure for performance metrics (Figure 2) is displayed.

```{r, fig.width = 10, fig.height = 6}
x_plots <- plot(x)
```

To save the figure for performance metrics to a PDF file:

```{r}
ggsave(x_plots$`Performance metrics`, filename = "output/compas_metrics.pdf", 
       width = 20, height = 12, units = "cm")
```

To format and compile other figures (Figure 3), and save to a PDF file:

```{r, fig.width=10, fig.height=13}
common_theme <- theme(legend.position = "bottom", 
                      legend.box.spacing = unit(0, "lines"))
p_roc <- x_plots$`ROC curves` + common_theme + 
  theme(axis.title.y = element_text(margin = margin(0, 5, 0, 0)))
p_calib <- x_plots$`Calibration curves` + common_theme
p_calib_large <- x_plots$`Calibration in the large` + common_theme +
  guides(
    fill = guide_legend(ncol = 2, title = "Group")
  )
p_score <- x_plots$`Boxplot of predictions` + common_theme + 
  guides(
    color = guide_legend(ncol = 2, title = "Group"),
    linetype = guide_legend(ncol = 2, title = "Label")
  )
p_nnp <- x_plots$`Number needed for true positive` + common_theme + 
  theme(axis.title.y = element_text(margin = margin(0, 5, 0, 0))) + 
  guides(
    color = guide_legend(ncol = 2, title = "Group")
  )
p_nnn <- x_plots$`Number needed for true negative` + common_theme + 
  theme(axis.title.y = element_text(margin = margin(0, 5, 0, 0))) + 
  guides(
    color = guide_legend(ncol = 2, title = "Group")
  )
p_all <- ggarrange(
  p_roc, p_calib, p_calib_large, p_score, p_nnp, p_nnn,
  labels = LETTERS[1:6], font.label = list(size = 14, face = "bold"),
  ncol = 2, nrow = 3,
  heights = c(6, 3.5, 4.5)#, align = "v"
)
p_all
ggsave(p_all, filename = "output/compas_other_fairness.pdf", 
       width = 23, height = 29.9, units = "cm")
```
